Lattice Boltzmann study of mass transfer for two-dimensional Bretherton/Taylor bubble train flow

### 2012-07-22 Text/introduction.tex

The understanding of mass transfer for the bubble train flow is not possible without deep understanding of hydrodynamic patterns.
There are several works studying the hydrodynamic properties of the bubble train flow, both
experimental \cite{kreutzer-pressure-drop,cerro-space,cerro-bubble-train} and numerical \cite{wang-non-circular,kuzmin-binary3d,giavedoni-numerical,heil-threedim}.
 For the flow of long bubbles between parallel plates chosen here as the study case, it is indicated that  there exists a vortex 
in the liquid slug for $Ca<0.7$, see Fig. \ref{fig:streamlines:tweaked:velocity}, and that the bubble shape is symmetric for low capillary 
numbers ($Ca<0.1$ \cite{cerro-bubble-train}) with the capillary number defined as:
 \beq
 Ca=\frac{\mu_{\mathrm{liq}} \ububble}{\gamma},
 \feq
 where $\mu_{\mathrm{liq}}$ is the liquid viscosity, $\ububble$ is the bubble velocity, and $\gamma$ is the interfacial tension.
 The fact that bubble shape for $Ca<0.1$ can be represented as two hemicircles and two planar interfaces with the vortex existing in the liquid slug has been 
utilized for analytical estimations of mass transfer properties.

@@

Despite their simplicity, such analytical expressions work well for flows with low
capillary numbers $Ca<0.1$ \cite{bercic-mass} where the bubble shape is
symmetrical and can approximated with good precision. Moreover, because of the
hydrodynamic pattern in the slug (i.e. presence of a vortex in slug), one can
estimate the time for a fluid batch to travel the entire circumference.   However,
with the increase of the capillary number the situation drastically changes -- the
symmetrical bubble shape is lost and the bubble shape resembles a bullet \cite{kuzmin-binary2d}.
For flows with $Ca>0.7$ there is also no vortex in the liquid slug anymore. The Higbie theory fails
to estimate the mass transfer from bubble caps, which explains the need to turn to numerical simulations where all
hydrodynamics fields as well as complex bubble shapes are taken into account.

Typical numerical studies of mass transfer
\cite{kreutzer-overview,vanbaten-circular} do not consider the simulation of
bubble shapes for $Ca>0.1$. The usual simulation of mass transfer is performed as
follows: 
\begin{description}
 \item[I] The bubble shape is calculated either
through analytical correlations \cite{bretherton} or experimental correlations
\cite{cerro-bubble-train} without directly resolving the bubbles shapes through
multiphase simulations.  The expressions for bubble shapes are available only
for flows with capillary number $Ca<0.1$.
  \item[II] Hydrodynamic fields
are then obtained by performing simulations of one-component flow around the
bubble by imposing the bubble velocity on the walls. Thus, the simulations are
performed in the reference frame moving with the bubble. A stress-free
condition is imposed at the bubble surface.
  \item[III]  The mass transfer
simulations are performed in the reference frame moving with the bubble. The
saturation concentration is imposed at the bubble surface. Only one unit cell
containing a single bubble is used for simulations. Periodic concentration boundary conditions are
utilized to determine the volumetric mass transfer coefficient, which is
calculated through the following equation \cite{vanbaten-circular}:
\begin{equation} \label{main:simulation:equation}
\vol=\frac{\mathrm{\overline{Flux}}}{\cbubble-\langle\cinoutlet\rangle}
\frac{\mathrm{bubble\ surface\ area}}{\mathrm{unit\ cell\ volume}},
\end{equation}
 where $\langle\cinoutlet(t)\rangle=\int{C \uinoutlet
\mathrm{d}A}/\int{\uinoutlet\mathrm{d}A}$ is the space-averaged inlet/outlet (periodic boundary conditions)
concentration as a function of time. Therefore, in terms of the mass transfer definition, $\langle\cinoutlet(t)\rangle$ plays a role of the 
characteristic concentration.
%This is the characteristic concentration in the definition, Eq. \ref{eq:main:definition}.  
The time-averaged
concentration flux ($\mathrm{\overline{Flux}}$) is calculated as the difference
between the overall average concentration in the whole domain
($\langle\coverall\rangle=\int_{V} C \mathrm{d}V /V$) at time $t_1$ and at time
$t_2$ divided by the time difference $t_2-t_1$. The agreement between numerical
simulations \cite{vanbaten-circular}, which formulation is presented above, and experimental correlations of \citet{bercic-mass}
% TODO: was good where?
was good. 
 \end{description}

@@

While it is clear that periodic boundary conditions can be employed for the calculation
 of hydrodynamic fields, the same does not apply to the mass transfer coefficient
simulations. Experimental correlations \cite{bercic-mass} show that the
concentration along the streamwise direction changes exponentially with distance.
Mass transfer simulations however, consider only one unit cell
using periodic boundary conditions, and thus the same concentration at the inlet
and at the outlet. The question how one unit cell simulations correspond to
experimental measurements where concentration differences are measured at
 distances of at least of a few unit cells \cite{bercic-mass}. In
other words one needs to understand how the discrete one unit cell simulation
corresponds to the continuous picture in experiments where one does not
distinguish discrete bubbles but takes measurements of concentration at different
locations.

Addressing situations for a rich number of hydrodynamic patterns, shapes, and
effects of bubble lengths, etc for bubble train flows, we feel that there is a need to examine 
carefully the strategies and assumptions behind the numerical calculations
of the mass transfer coefficient. We aim at establishing clear
procedures as to how properly obtain the mass transfer coefficient via a study
of different boundary conditions and different definitions of the characteristic concentration.
The case we want to examine is a two-dimensional bubble train flow between parallel plates.
We address the following issues:
 \begin{description}
 \item[I] Applicability of periodic boundary
conditions to determine the mass transfer coefficient when the vortex in the slug
disappears, i.e. $Ca>0.7$.
 \item[II] Validity of the inlet/outlet-averaged or domain-averaged concentrations as
characteristic concentrations in the definition of the mass transfer
coefficient.
 \item[III] Transition of the continuous experimental picture to
numerical simulations with a few unit cells, the issue of correspondence
between space averages (simulations \cite{vanbaten-circular}) and time averages (experiment).
% \item[IV] Influence of the hydrodynamic pattern in flows with $Ca>0.7$ on the mass transfer.
\end{description}
In addition, at the end of the manuscript we present results of the dependence of the volumetric
mass transfer coefficient on the Peclet number that we compare with analytical \cite{irandoust} and experimental correlations
\cite{yue-mass}.  The thorough determination of the mass transfer coefficient and associated Sherwood number as a
function of other non-dimensional parameters such as gas holdup, bubble/slug
lenghts, and the capillary number is left for future.
\begin{figure}[htb!]
\includegraphics[width=\textwidth]{Figures/benchmark_hydro.eps}
\caption{Simplified sketch of the bubble motion. Using periodic
conditions for the velocity field is natural, but needs evaluation for mass
transfer.
\label{fig:benchmark:hydro}}
\end{figure}

@@

\subsection{Average concentration results}
\label{main:results:periodic}
In this section we will examine the case where the volume averaged concentration over
 time is used as the characteristic concentration. To calculate the volumetric mass transfer coefficient we used Eq. \ref{theor:one:concentration:time}.
%\beqal
%&\vol t \frac{\ububble}{\ugas+\uliq}=\ln\frac{\cstar}{\cstar-\langle C(t) \rangle}\\
%&\vol \frac{\lunit}{\ugas+\uliq}=\frac{\lunit}{\ububble t} \ln \frac{C^*}{C^*-\langle C(t)
%\rangle},\\
%\feqal
%where $\langle C(t) \rangle$ is the average concentration in the liquid domain.
Results for the coefficient $\vol \frac{\ububble}{\ugas+\uliq}$ are shown in Fig.
\ref{fig:aver:conc:different:capillaries:time} for different Peclet numbers and velocity scalings
indicated in Table \ref{table:scaling:peclet}. When the average concentration gets
close to $\cstar=1$ then Eq. \ref{theor:one:concentration:time} gives inadequate results due to
 accuracy of the logarithmic function evaluation. 
 %Fig. \ref{fig:aver:conc:different:capillaries:time} presents
%the volumetric mass transfer coefficient as a function of number of iterations. 
Due to velocity scaling
each simulation has a different physical time step. Thus, we normalized time such that it represents a number of cell
units which tracer will pass with the given bubble velocity, i.e. 
$N_{\text{cell\,units}}=\frac{\text{scale}\cdot \ububble\cdot N_{\mathrm{iter}}}{\lunit}$.  Fig.
\ref{fig:aver:conc:different:capillaries} shows the volumetric mass transfer
dependency against the distance in unit cell length. One can see in Table
\ref{table:steady:state:average} that for different Peclet numbers different time (number of unit
cells) is required to achieve  steady state. For example, for
larger Peclet numbers fewer iterations are required to achieve the steady state condition.

>>>>

The understanding of mass transfer for the bubble train flow is not possible without a deep understanding of hydrodynamic patterns.
There are several works studying the hydrodynamic properties of the bubble train flow, both
experimental \cite{kreutzer-pressure-drop,cerro-space,cerro-bubble-train} and numerical \cite{wang-non-circular,kuzmin-binary3d,giavedoni-numerical,heil-threedim}.
 For the flow of long bubbles between parallel plates chosen here as the study case, it is indicated that  there exists a vortex 
in the liquid slug for $Ca<0.7$, see Fig. \ref{fig:streamlines:tweaked:velocity}, and that the bubble shape is symmetric for low capillary 
numbers ($Ca<0.1$ \cite{cerro-bubble-train}) with the capillary number defined as:
 \beq
 Ca=\frac{\mu_{\mathrm{liq}} \ububble}{\gamma},
 \feq
 where $\mu_{\mathrm{liq}}$ is the liquid viscosity, $\ububble$ is the bubble velocity, and $\gamma$ is the interfacial tension.
 The fact that bubble shape for $Ca<0.1$ can be represented as two hemicircles and two planar interfaces with the vortex existing in the liquid slug has been 
utilized for analytical estimations of mass transfer properties.

@@

Despite their simplicity, such analytical expressions work well for flows with low
capillary numbers $Ca<0.1$ \cite{bercic-mass} where the bubble shape is
symmetrical and can be approximated with good precision. Moreover, because of the
hydrodynamic pattern in the slug (i.e. presence of a vortex in slug), one can
estimate the time for a fluid batch to travel the whole circumference.   However,
with the increase of the capillary number the situation changes significantly -- the
symmetrical bubble shape is lost and the bubble shape resembles a bullet \cite{kuzmin-binary2d}.
For flows with $Ca>0.7$ there is also no vortex in the liquid slug. In this case the Higbie theory fails
to estimate the contribution from bubble caps, which explains the need to turn to numerical simulations where all
hydrodynamic fields as well as complex bubble shapes are taken into account.

Typical numerical studies of mass transfer
\cite{kreutzer-overview,vanbaten-circular} do not consider the simulation of
bubble shapes for $Ca>0.1$. The usual simulation of mass transfer is performed as
follows: 
\begin{description}
 \item[I] The bubble shape is calculated either
through analytical correlations \cite{bretherton} or experimental correlations
\cite{cerro-bubble-train} without directly resolving bubble shapes through
multiphase simulations.  The expressions for bubble shapes are available only
for flows with capillary number $Ca<0.1$.
  \item[II] Hydrodynamic fields
are then obtained by performing simulations of one-component flow around the
bubble by imposing the bubble velocity on the walls. Thus, the simulations are
performed in the reference frame moving with the bubble. A stress-free
condition is imposed at the bubble surface.
  \item[III]  The mass transfer
simulations are performed in the reference frame moving with the bubble. The
saturation concentration is imposed at the bubble surface. Only one unit cell
containing a single bubble is used for simulations. Periodic concentration boundary conditions are
utilized to determine the volumetric mass transfer coefficient, which is
calculated through the following equation \cite{vanbaten-circular}:
\begin{equation} \label{main:simulation:equation}
\vol=\frac{\mathrm{\overline{Flux}}}{\cbubble-\langle\cinoutlet\rangle}
\frac{\mathrm{bubble\ surface\ area}}{\mathrm{unit\ cell\ volume}},
\end{equation}
 where $\langle\cinoutlet(t)\rangle=\int{C \uinoutlet
\mathrm{d}A}/\int{\uinoutlet\mathrm{d}A}$ is the space-averaged inlet/outlet (periodic boundary conditions)
concentration as a function of time. Therefore, in terms of the mass transfer definition, $\langle\cinoutlet(t)\rangle$ plays a role of the 
characteristic concentration.
%This is the characteristic concentration in the definition, Eq. \ref{eq:main:definition}.  
The time-averaged
concentration flux ($\mathrm{\overline{Flux}}$) is calculated as the difference
between the overall average concentration in the whole domain
($\langle\coverall\rangle=\int_{V} C \mathrm{d}V /V$) at time $t_1$ and at time
$t_2$ divided by the time difference $t_2-t_1$. The agreement between numerical
simulations \cite{vanbaten-circular}, which formulation is presented above, and experimental correlations of \citet{bercic-mass}
% TODO: was good where?
was good. 
 \end{description}

@@

While it is clear that periodic boundary conditions can be employed for the calculation
 of hydrodynamic fields, the same does not apply to the mass transfer coefficient
simulations. Experimental correlations \cite{bercic-mass} show that the
concentration along the streamwise direction changes exponentially with distance.
Mass transfer simulations however, are made only for one unit cell
using periodic boundary conditions with the same concentration at the inlet
and at the outlet. The question is how one unit cell simulation corresponds to
experimental measurements arises where concentration difference is measured at
the distances of at least a few unit cells \cite{bercic-mass}. In
other words one needs to understand how the discrete one unit cell simulation
corresponds to the continuous picture in experiments where one does not
distinguish discrete bubbles but takes measurements of concentration at different
locations.

Addressing situations for a rich number of hydrodynamic patterns, shapes, and
effects of bubble lengths, etc for bubble train flows, we feel that there is a need to examine 
carefully the strategies and assumptions behind the numerical calculations
of the mass transfer coefficient. We aim at establishing clear
procedures as to how properly obtain the mass transfer coefficient via a study
of different boundary conditions and different definitions of the characteristic concentration.
The case we want to examine is a two-dimensional bubble train flow between parallel plates.
We address the following issues:
 \begin{description}
 \item[I] Applicability of periodic boundary
conditions to determine the mass transfer coefficient when the vortex in the slug
disappears, i.e. $Ca>0.7$.
 \item[II] Validity of the inlet/outlet-averaged or domain-averaged concentrations as
characteristic concentrations in the definition of the mass transfer
coefficient.
 \item[III] Translation of the continuous experimental picture to
numerical simulations of a few unit cells, the issue of correspondence
between space averages (simulations \cite{vanbaten-circular}) and time averages (experiment).
% \item[IV] Influence of the hydrodynamic pattern in flows with $Ca>0.7$ on the mass transfer.
\end{description}
In addition, at the end of the manuscript we present results of the dependence of the volumetric
mass transfer coefficient on the Peclet number that we compare with analytical \cite{irandoust} and experimental correlations
\cite{yue-mass}.  The thorough determination of the mass transfer coefficient and associated Sherwood number as a
function of other non-dimensional parameters such as gas holdup, bubble/slug
lenghts, and the capillary number is left for future.
\begin{figure}[htb!]
\includegraphics[width=\textwidth]{Figures/benchmark_hydro.eps}
\caption{Simplified sketch of the bubble motion. Using periodic
conditions for the velocity field is natural, but needs evaluation for mass
transfer.
\label{fig:benchmark:hydro}}
\end{figure}

@@

\subsection{Average concentration results}
\label{main:results:periodic}
In this section we will examine the case where the volume averaged concentration over
 time is used as the characteristic concentration. To calculate the volumetric mass transfer coefficient we used Eq. \ref{theor:one:concentration:time}.
%\beqal
%&\vol t \frac{\ububble}{\ugas+\uliq}=\ln\frac{\cstar}{\cstar-\langle C(t) \rangle}\\
%&\vol \frac{\lunit}{\ugas+\uliq}=\frac{\lunit}{\ububble t} \ln \frac{C^*}{C^*-\langle C(t)
%\rangle},\\
%\feqal
%where $\langle C(t) \rangle$ is the average concentration in the liquid domain.
Results for the coefficient $\vol \frac{\ububble}{\ugas+\uliq}$ are shown in Fig.
\ref{fig:aver:conc:different:capillaries} for different Peclet numbers and velocity scalings
indicated in Table \ref{table:scaling:peclet}. When the average concentration gets
close to $\cstar=1$ then Eq. \ref{theor:one:concentration:time} gives inadequate results due to
 accuracy of the logarithmic function evaluation. 
 %Fig. \ref{fig:aver:conc:different:capillaries:time} presents
%the volumetric mass transfer coefficient as a function of number of iterations. 
Due to velocity scaling
each simulation has a different physical time step. Thus, we normalized time such that it represents a number of cell
units which tracer will pass with the given bubble velocity, i.e. 
$N_{\text{cell\,units}}=\frac{\text{scale}\cdot \ububble\cdot N_{\mathrm{iter}}}{\lunit}$.  Fig.
\ref{fig:aver:conc:different:capillaries} shows the volumetric mass transfer
dependency against the distance in unit cell length. One can see in Table
\ref{table:steady:state:average} that for different Peclet numbers different time (number of unit
cells) is required to achieve  steady state. For example, for
larger Peclet numbers fewer iterations are required to achieve the steady state condition.
